---
title: "NFL Time Series Prediction"
author: "Angelica"
date: "5/16/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	fig.height = 5,
	fig.width = 8,
	message = FALSE,
	warning = FALSE,
	results = "hide"
)
library(tidyverse) #data structure
library(ggplot2) #visualization
library(forecast) #prediction
```


## Introduction

The National football league (NFL) is a 6.5-billion-dollar brand, gaining international popularity by the year. In the last year alone, (2020 – 2021), the NFL grew in popularity by 14%. Additionally, the NFL adds 270 million to the gambling industry. Needless to say, understanding the analytics of football could score you a heavy paycheck. 

The NFL consists of two divisions, the American Football Conference (AFC) and the National Football Conference (NFC). Currently, each division has 16 teams, for a total of 32 teams. For this project, I recreated the NFL data from the past 30 years (1991-2021), and the idea is to build a prediction/forecasting model of the number of wins for a particular team of our choice within the two NFL divisions and then compare the team’s forecast predictions. 
For more information about NFL teams, see this link here; <https://www.nfl.com/teams/>. 


*Data source: <https://www.nfl.com/standings/>*

*Notes:*

This is only regular season wins and losses, playoff games are not included. Thus, the forecasting will be on number of wins in a regular season. 


## Dataset Description

There are 13 variables: 

----------------------------------------------------------
Variable    Description                        Data Type
---------- ---------------------------------- ------------
Tm          American Football Team             chr                        

W           Games Won                          num

L           Games Lost                         num

W-L%        Win-Loss Percentage of Team        num

PF          Points Scored by Team              num

PA          Points Scored by Opposition        num

PD          Points Differential(PF - PA)       num

MoV         Margin of Victory                  num

SoS         Average quality of opponent        num

SRS         Simple Rating System(MoV+SoS)      num

OSRS        Offensive SRS                      num

DSRS        Defensive SRS                      num

Year        Year of the NFL season             num

-------------------------------------------------------


## Exploratory Analysis


Starting with an exploratory analysis of the overall wins of each team in their respective NFL division. The time series plots from this analysis let us proceed with the team selection.


```{r}
# import data 
football <- readxl::read_xlsx("Football_Project.xlsx")
#head(football)
```


```{r}
# create a time series object
football_ts <- ts(data = football[,5], start = 2001, end = 2021, frequency = 1)
```


## Overall Wins on all teams in AFC football division


```{r}
# Plot of all teams in AFC football division
ggplot(football, aes(x = Year, y = W)) +
  geom_line(aes(color = Tm, group = Tm)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(y = "Wins") +
  facet_wrap(~ Tm) 
```


## Overall wins on all teams in NFC football division


```{r}
NFL <- readxl::read_xlsx("Football_Project2.xlsx")
#NFL
NFL2 <- NFL[order(NFL$Tm),]
#NFL2


P <- ggplot(NFL2, aes(x=Year,y=W,colour=Tm,group=Tm)) + geom_line()

P2 <- P + facet_wrap(~Tm) +                                                   
  theme(strip.text.x = element_text(size = 10),
        axis.text.x = element_text(colour="grey20", 
                                   size=12, angle=90, hjust=.5, vjust=.5))

P2
```


## Predictive Strategy


**Prediction Problem:** Create a model to predict the number of wins for a particular NFL team in the next 2 years (2022-2023).


**Description of the Strategy**

1. Plot the dataset and remove trend or seasonality if it is needed.

2. fit the best model implementing different forecasting methods.

3. Residual diagnosis to check best model assumptions.

4. Forecasting the number of wins for the next 2 years based on one NFL team.


## Analyze Time Series on One Specific Team


Based on the overall wins on the NFL teams, the New England Patriots is the team for time series analyses. The main goal is to build the best model and create a forecasting comparison with the final result.


## Transformation of the New England Patriots' Win data


```{r}
# filter a team from the dataset
football_team <- filter(.data = football, Tm == "New England Patriots")

# create a time series object
football_team_ts <- ts(data = football_team$W, start = 1991, end = 2021, 
                       frequency = 1)

# plot of the original series
plot(football_team_ts, xlab="Year", ylab = "", 
     main = "Wins of New England Patriots Team")

# plot of the logarithms of the series
football_log <- log(football_team_ts)
plot(football_log, xlab="Year", ylab = "", 
     main = "Log Wins of New England Patriots Team")
```

We can see a moderate increasing trend from 1992 to 2006. Log is an appropriate transformation to stabilize the data from the increasing trend over that period. 


```{r}
#plot of the first difference of the logged series.
diff1 <- c(NA, diff(log(football_team_ts)))
diff1 <- ts(diff1, start = c(1991,1), deltat = 1)

plot(diff1, xlab = "Year", ylab = "", main = "First order differenced series")
```


It appears that trend have been removed after applying First order differenced series. The differential order makes the time series stationary.


```{r}
## ACF & PACF plots
par(mfrow = c(1, 2)) 
acf(football_team_ts)#, lag.max = 15)
pacf(football_team_ts)#, lag.max = 15)

auto.arima(football_team_ts)

par(mfrow = c(1, 2)) 
acf(diff1, lag.max = 12, na.action = na.pass, main = "ACF for differenced series")
pacf(diff1, lag.max = 12, na.action = na.pass, main = "PACF for differenced series")

auto.arima(diff1)
```


Based on the AIC resulting from the auto.arima function, ARIMA(0,1,1)  could be the best fit model for this particular dataset. Next step, try a few models to confirm this assumption.


## Fitting the best ARIMA model


## AR(1)

```{r, results='markup'}
fit_ar1 <- arima(football_log, order = c(p = 1, d = 0, q = 0))
fit_ar1
```


## AR(2) model

```{r, results='markup'}
fit_ar2 <- arima(football_log, order = c(p = 2, d = 0, q = 0))
fit_ar2
```

## ARIMA(0,1,1)

```{r, results='markup'}
fit_arima <- arima(football_log, order = c(p = 0, d = 1, q = 1))
fit_arima
```


## ARIMA(1,1,1)

```{r, results='markup'}
fit_arima1 <- arima(football_log, order = c(p = 1, d = 1, q = 1))
fit_arima1
```



According to the above models, we can confirm ARIMA(0,1,1) is an appropriate model since it gives the smallest AIC value.


## Residuals Diagnostic


```{r}
## Fitting residuals 
par(mfrow = c(1, 2)) 
res <- fit_arima$residuals 

# ACF & PACF plots
acf(res, lag.max = 12, na.action = na.pass) 
pacf(res, lag.max = 12, na.action = na.pass)

# diagnostic
tsdiag(fit_arima)
```

```{r}
# check normality of the residuals
qqnorm(res) 
qqline(res)
```


The sample ACF and PACF plot on the diagnostic suggest the residuals follow white noise, and QQ plot indicates the normality assumption doesn’t seem to be violated. Overall, the fitted model looks good.


## Forecasting


```{r}
# prediction of the time series
pred <- predict(fit_arima, n.ahead = 2)

# prediction plot
plot(football_team_ts, xlim = c(1991, 2024), ylim = c(2,20), 
     ylab = "", xlab = "Year", 
     main = "New England Patriots Team and Forecast the Next 2 Years Wins")

## forecasted values 
lines(exp(pred$pred), col = "red") 

## 95% forecasting limits 
lines(exp(pred$pred-2*pred$se),col='blue') 
lines(exp(pred$pred+2*pred$se),col='blue')

```


The plotted series above is based on the `predict()` function in R, and contains 2 arguments. In the first argument, the best fit ARIMA model, and in the second argument, the number of periods for forecasting, in this case, the following 2 years. Also, the blue lines represent the 95% forecasting limits and the red line the forecasting values for the 2 years ahead.


```{r}
## Simple exponential smoothing
forecast_team <- ses(y = football_log, h = 2) 
plot(forecast_team)
```


The visualization above, the `ses()` function in R is implemented. It stands for `simple exponential smoothing`; and the arguments `y` and `h` indicate the time series data and the number of periods for forecasting.

The simple exponential smoothing model is considered to use on stationary series. The forecasting based on this model is performing as expected and looks pretty similar to the prediction model.



## Conclusion


Based on the two forecasting visualizations, we can appreciate `New England Patriots` team has a 95% predicted interval between 5 wins to 15 wins, with the most likely outcome being 10 wins for the next 2 years (2022-2023). As we can see, using different methods to test the NFL data lead us to fit the best forecasting model and finally plot it for a better understanding. In the future, analyzing a little further the team (i.e, players) would be good option to obtain a better forecasting prediction. 



## Limitations


Creating a prediction model can lead us to perform different models, but sometimes is difficult when it is the right time to stop while we are building the best model. Also, in the NFL dataset, we have to think about the team as a whole. 

- if the team will stay the same for the next 2 years (i.e., players, coach, ...)
- if the best players receive any injuries during the next 2 regular seasons. That can affect their overall performance.



## Appendix

```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```